---
title: "Exploring protected areas and dam locations in Nepal"
author: "Aakriti Poudel"
date: "February 2026"
format:
  pdf:
    code-fold: false
    toc: true
  html:
    code-fold: false
    toc: true
execute:
  warning: false
  message: false
editor: source
editor_options:
  chunk_output_type: inline
---

## Overview

This document explores the spatial relationship between protected areas, current dams and proposed future dams within Nepal's river basin. The goal is to visually identify where future dams development might overlap with socially and ecologically sensitive areas.

```{r}
#| echo: false
#| output: false

# Check the current working directory
getwd()

```

## Set up

Load all necessary packages

```{r}
#| label: load-packages

# File path management
library(here)

# Data cleaning and wrangling
library(janitor)
library(tidyverse)

# Spatial data handling
library(sf)

# Interactive and static mapping
library(tmap)
library(leaflet)

```

### Check the protected areas geodatabase

The WDPA geodatabase contains three layers: polygons, points and source metadata. We need the polygon layer for our spatial analysis.

```{r}
#| label: check-layers

# Check what layers are inside the data set
st_layers("../../../../../capstone/netzerohydro/data/raw/WDPA_WDOECM_Jan2026_Public/world_database_protected_areas.gdb")

```

## Read all datasets

```{r}
#| label: read-data
#| output: false

# Read the polygon layer for protected areas
protected_areas_polygon <- st_read(
  "../../../../../capstone/netzerohydro/data/raw/WDPA_WDOECM_Jan2026_Public/world_database_protected_areas.gdb",
  layer = "WDPA_WDOECM_poly_Jan2026_e05cdb98a28bc5c1f76cc4919558750e5cb5dfe8ebdf74a197e38cf76ee489b7") %>%
  clean_names()

# Read the current dams dataset
current_dams <- st_read("../../../../../capstone/netzerohydro/data/raw/GRanD_Version_1_3") %>% clean_names()

# Read the current dams dataset
future_dams <- read_csv("../../../../../capstone/netzerohydro/data/raw/FHReD_2015_future_dams/FHReD_2015_future_dams_Zarfl_et_al_beta_version.csv") %>%
  clean_names()

# Read watershed boundaries at basin level 3
basin_level3 <- st_read("../../../../../capstone/netzerohydro/data/raw/hybas_as_lev03_v1c") %>% clean_names()

```

## Explore all datasets

### Convert future dams to a spatial object

The FHReD dataset is a CSV file, so it does not have any spatial information. We convert it into a spatial object by pointing to its longitude and latitude columns and assigning the WGS 84 coordinate reference system (CRS).

```{r}
#| label: convert-future-dams

# Convert the CSV to a spatial (sf) object using its coordinate columns
future_dams_sf <- future_dams %>%
  st_as_sf(coords = c("lon_cleaned", "lat_cleaned"), crs = 4326)

```

### Verify coordinate reference systems (CRS)

Before combining or overlaying spatial datasets, we need to make sure they all use the same coordinate system. We check the CRS of each dataset below.

```{r}
#| label: check-crs
#| output: false

# Check CRS for each layer
st_crs(protected_areas_polygon)
st_crs(current_dams)
st_crs(future_dams_sf)
st_crs(basin_level3)

```

All four datasets use **WGS 84 (EPSG:4326)**, so no coordinate transformation is needed.

## Filter to Nepal basin

We use the HydroBASINS level 3 dataset to isolate Nepal's watershed. This basin-based approach is more appropriate than using political boundaries because our analysis focuses on river connectivity and watershed level impacts.

```{r}
#| label: filter-nepal-basin

# Filter to Nepal basin
nepal_basin <- basin_level3 %>% 
  filter(hybas_id == 4030025450)

```

### Crop datasets to Nepal

Each global dataset needs to be clipped to the Nepal basin boundary. Here, we use two approach for efficiency. **First**, we crop to the bounding box, a quick rectangular filter. **Second**, we perform a precise intersection with the basin polygon.

We temporarily turn off spherical geometry (S2) because these datasets contain minor geometry errors that cause S2 to fail. For a small area, planar geometry is accurate enough and avoids these issues.

*S2 is a system for working with geographic data on a sphere. It is is used by the sf package in R by default. S2 is more accurate, but it's also stricter — it will reject geometries with even tiny errors like a line crossing itself or a duplicated point. That's why we turned it off temporarily when working with the WDPA and GRanD datasets, which had minor geometry imperfections. For a small region like Nepal, the difference in accuracy between spherical and flat calculations is negligible.*

### Crop protected areas

```{r}
#| label: crop-protected-areas

# Turn off spherical geometry to avoid errors from minor geometry issues
sf_use_s2(FALSE)

# Crop to bounding box first, fix any invalid geometries, then intersect precisely
nepal_protected <- protected_areas_polygon %>%
  st_crop(st_bbox(nepal_basin)) %>%
  st_make_valid() %>%
  st_intersection(nepal_basin)

# Turn spherical geometry back on
sf_use_s2(TRUE)
```

### Crop current dams

```{r}
#| label: crop-current-dams

# Turn off spherical geometry to avoid errors from minor geometry issues
sf_use_s2(FALSE)

# Crop to bounding box first, fix any invalid geometries, then intersect precisely
nepal_dams <- current_dams %>%
  st_crop(st_bbox(nepal_basin)) %>%
  st_make_valid() %>%
  st_intersection(nepal_basin)

# Turn spherical geometry back on
sf_use_s2(TRUE)

```

### Crop future dams

```{r}
#| label: crop-future-dams

# Turn off spherical geometry to avoid errors from minor geometry issues
sf_use_s2(FALSE)

# Crop to bounding box first, fix any invalid geometries, then intersect precisely
nepal_future_dams <- future_dams_sf %>%
  st_crop(st_bbox(nepal_basin)) %>%
  st_make_valid() %>%
  st_intersection(nepal_basin)

# Turn spherical geometry back on
sf_use_s2(TRUE)

```

## Clean text encoding

Some datasets contain special characters that are not valid UTF-8, which can cause errors when rendering interactive maps. We clean all text columns by converting them to UTF-8 and removing any invalid characters.

*UTF-8 is a standard way computers store and display text. It covers English letters, numbers and characters from other languages like Nepali, Chinese, or Spanish. Some of our datasets (especially WDPA and FHReD) contain place names or descriptions with special characters that don't follow this standard. These characters look fine in a table, but when tmap tries to build an interactive Leaflet map, it converts everything to a web-friendly format that requires valid UTF-8. The `iconv()` function goes through every text column, converts characters to proper UTF-8, removes any that can't be converted. This way, we might lose the occasional unusual character, but the maps render without errors.*

```{r}
#| label: fix-encoding

# Convert all character columns to valid UTF-8 encoding
nepal_protected <- nepal_protected %>%
  # `sub` substitutes invalid characters with nothing
  mutate(across(where(is.character), ~iconv(., to = "UTF-8", sub = "")))

nepal_future_dams <- nepal_future_dams %>%
  mutate(across(where(is.character), ~iconv(., to = "UTF-8", sub = "")))
```

## Set map mode

We switch `tmap` to interactive mode, which creates zoomable, clickable maps powered by Leaflet. This makes it easy to explore the data and click on individual features to see their attributes.

```{r}
#| label: set-tmap-mode

# Use interactive (Leaflet-based) maps
tmap_mode("view")
```

### Map 1: Current dams and protected areas

This map shows existing dams (blue dots) overlaid on protected areas (green polygons) within the Nepal basin. It helps us see how many current dams are located near or within ecologically protected zones.

```{r}
#| label: map-current-dams

tm_shape(nepal_basin) +
  tm_polygons(fill = "grey90", col = "grey60") +
tm_shape(nepal_protected) +
  tm_polygons(fill = "darkgreen", fill_alpha = 0.2, col = "darkgreen") +
tm_shape(nepal_dams) +
  tm_dots(fill = "deepskyblue3", size = 0.25) +
tm_title("Current Dams and Protected Areas in Nepal Basin")
```

### Map 2: Future dams and protected areas

This map shows planned future dams (violet red dots) against protected areas. It highlights where proposed hydropower development could conflict with conservation goals.

```{r}
#| label: map-future-dams

tm_shape(nepal_basin) +
  tm_polygons(fill = "grey90", col = "grey60") +
tm_shape(nepal_protected) +
  tm_polygons(fill = "darkgreen", fill_alpha = 0.2, col = "darkgreen") +
tm_shape(nepal_future_dams) +
  tm_dots(fill = "violetred4", size = 0.25) +
tm_title("Future Dams and Protected Areas in Nepal Basin")
```

### Map 3: Current and future dams together with protected area

This combined map shows both existing dams (blue dots) and planned future dams (violet red dots) alongside protected areas. Current dams are plotted last so they remain visible on top of the future dam layer. This view helps us compare where dams already exist versus where new ones are planned, and how both relate to protected areas.

```{r}
#| label: map-protected-areas-dams

tm_shape(nepal_basin) +
  tm_polygons(fill = "grey90", col = "grey60") +
tm_shape(nepal_protected) +
  tm_polygons(fill = "darkgreen", fill_alpha = 0.2, col = "darkgreen") +
tm_shape(nepal_future_dams) +
  tm_dots(fill = "violetred4", size = 0.25) +
tm_shape(nepal_dams) +
  tm_dots(fill = "deepskyblue", size = 0.25) +
tm_title("Current & Future Dams with Protected Areas in Nepal Basin")
```

## Global analysis: Flag current and future dams in protected areas

In the Nepal exploration, we visually looked at where dams and protected areas overlap. Now, we scale this analysis for the global data set. Instead of cropping and mapping, we use `spatial join` to attach protected area information to each dam. This tells us, for every dam in the world, whether it falls inside a protected area, how many protected areas it overlaps, and what type of protection those areas have (IUCN category).

The key difference from Nepal workflow:

- `st_intersection` (used for Nepal workflow) clips and cuts geometries to a boundary, useful for mapping.
- `st_join` (used for global analysis) keeps dam points intact and simply looks up whether each dam falls inside a protected area, useful for data analysis.

## Prepare the protected areas dataset

The global WDPA data set has 292,731 polygons. Before running the spatial join, we fix any geometry errors in the dataset. We do this once and reuse it for both current and future dams.

### Flag current dams (GRanD)

For each of the 7,424 dams in the GRanD database, we check whether it falls inside any protected area. A spatial join (`st_join`) pairs each dam with any protected area polygon it touches. Since a single dam can overlap with more than one protected area, we then group by the dam's unique ID and collapse the results into one row per dam.

We turn off spherical geometry (S2) for the join. This makes the operation faster and more forgiving of minor geometry errors in the WDPA dataset, removing the need to run `st_make_valid()` on all 292,731 protected area polygons upfront.

The output has three new columns:

- `in_protected_area`: TRUE if the dam is inside at least one protected area, FALSE otherwise.
- `n_protected_areas`: The number of protected areas the dam overlaps.
- `iucn_categories`: A list of the IUCN categories of those protected areas (e.g., "II, IV").
```{r}
#| label: flag-current-dams

# Turn off spherical geometry — faster and avoids geometry errors in WDPA
sf_use_s2(FALSE)

# Step 1: Spatial join — attach protected area info to each current dam
# left = TRUE keeps all dams, even those not in any protected area
current_dams_joined <- current_dams %>%
  st_make_valid() %>%
  st_join(protected_areas_polygon, join = st_intersects, left = TRUE)

# Step 2: Summarise to one row per dam
# A dam may overlap multiple protected areas, creating duplicate rows
# We collapse these into a single row per dam with summary columns
current_dams_flagged <- current_dams_joined %>%
  mutate(in_protected_area = !is.na(site_id)) %>%
  group_by(grand_id) %>%
  summarise(
    # TRUE if the dam falls in at least one protected area
    in_protected_area = any(in_protected_area),
    # Count of how many protected areas the dam overlaps
    n_protected_areas = sum(in_protected_area),
    # Comma-separated list of IUCN categories
    iucn_categories = paste(unique(na.omit(iucn_cat)), collapse = ", "),
    .groups = "drop"
  )

# Turn spherical geometry back on
sf_use_s2(TRUE)
```

### Flag future dams (FHReD)

We repeat the same process for the 3,700+ planned future dams in the FHReD database. These are the dams our project will ultimately assess for suitability, so this protected area flag becomes one component of our "fatal flaw" analysis.

> **Note:** Check `colnames(future_dams)` to confirm the unique ID column for the FHReD dataset. The code below uses `dam_name` — replace it with the correct column if different.

```{r}
#| label: check-future-dam-columns

# Check column names to identify the unique ID for future dams
colnames(future_dams)

```


```{r}
#| label: flag-future-dams

# Turn off spherical geometry
sf_use_s2(FALSE)

# Step 1: Spatial join — attach protected area info to each future dam
# left = TRUE keeps all dams, even those not in any protected area
future_dams_joined <- future_dams_sf %>%
  st_make_valid() %>%
  st_join(protected_areas_polygon, join = st_intersects, left = TRUE)

# Step 2: Summarise to one row per dam
future_dams_flagged <- future_dams_joined %>%
  mutate(in_protected_area = !is.na(site_id)) %>%
  group_by(dam_id) %>%
  summarise(
    # TRUE if the dam falls in at least one protected area
    in_protected_area = any(in_protected_area),
    # Count of how many protected areas the dam overlaps
    n_protected_areas = sum(in_protected_area),
    # Comma-separated list of IUCN categories
    iucn_categories = paste(unique(na.omit(iucn_cat)), collapse = ", "),
    .groups = "drop"
  )

# Turn spherical geometry back on
sf_use_s2(TRUE)

```

### Summary: How many dams fall in protected areas?

A quick count to see how many current and future dams overlap with at least one protected area globally.
```{r}
#| label: summary-current-dams

# Count of current dams inside vs outside protected areas
current_dams_flagged %>%
  st_drop_geometry() %>%
  count(in_protected_area) %>%
  mutate(percent = round(n / sum(n) * 100, 1))
```


```{r}
#| label: summary-future-dams

# Count of future dams inside vs outside protected areas
future_dams_flagged %>%
  st_drop_geometry() %>%
  count(in_protected_area) %>%
  mutate(percent = round(n / sum(n) * 100, 1))
```

### Preview the flagged datasets
```{r}
#| label: preview-current-flagged

# Preview the first few rows of flagged current dams
current_dams_flagged %>%
  st_drop_geometry() %>%
  head(10)
```


```{r}
#| label: preview-future-flagged

# Preview the first few rows of flagged future dams
future_dams_flagged %>%
  st_drop_geometry() %>%
  head(10)
```
